Supplementary Information

### 2020-06-18 doc/manuscript/MEE_manuscript.Rmd
several small updates

For each year in the tree-ring records, we reconstructed $DBH$, as detailed in Appendix S1. 
In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings (DENDORO_REF), $DBH$ was reconstructed from the inside out. 
In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Table S2).

@@

Climate and environmental driver data were obtained from global databases. Monthly climate data for 1901-20XX was obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. Variables considered here included mean, minimum, and maximum temperatures; daily temperture range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET). Atmospheric CO~2~... (**[ISSUE #16 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/16): obtain data**). Atmospheric deposition... (**[ISSUE #17 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/17): obtain data**). Analyses were, necessarily, limited ot the time frame over which these were available.

@@

We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each of the three climate variable groups. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
Within *climwin*, we specified a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
If PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
If it came out as the best variable in one but not both of these groups, it was dropped from the analysis, on the logic that it was an inferior predictor to a related variable.
We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.

To check that this process identified similar climate variables to what would be identified using traditional methods, ... (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35): do a standardized analysis similar to Helcoski et al. for each site? or just compare with previous studies for each site?**)
(**[ISSUE #34 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/34): check that we get similar results with $\Delta r$ and $\Delta AGB$**)

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Fig. 2).**

>>>>

For each year in the tree-ring records, we reconstructed $DBH$, as detailed in *Appendix S1*. 
In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings (DENDORO_REF), $DBH$ was reconstructed from the inside out. 
In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Table S2).

@@

Climate and environmental driver data were obtained from global databases. Monthly climate data for 1901-20XX was obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. Variables considered here included mean, minimum, and maximum temperatures; daily temperture range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET). In the CRU database, gaps are filled with monthly means... (**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**). Atmospheric CO~2~... (**[ISSUE #16 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/16): obtain data**). Atmospheric deposition... (**[ISSUE #17 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/17): obtain data**). Analyses were, necessarily, limited ot the time frame over which these were available.

@@

We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each of the three climate variable groups. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods, as detailed in Appendix S2. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))

Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
If PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
If it came out as the best variable in one but not both of these groups, it was dropped from the analysis, on the logic that it was an inferior predictor to a related variable.
We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.

@@

**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. 2, S#; Appendix S2).**

### 2020-06-18 doc/manuscript/MEE_manuscript_SI.Rmd
update SI
several small updates

DBH can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $\Delta r$ from the inside out). 
We generally gave precedence to the outside-in approach.
Specifically, when $DBH$ was taken at the time of coring,  
At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used. 
If before, ... 
If after...
For all outside-in reconstructions, if a negative $DBH$ was predicted...

In either case we need bark thickness--ideally allometries. This is especially critical for thick-barked species

@@

>>>>

DBH can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $\Delta r$ from the inside out). 
We generally gave precedence to the outside-in approach.
Specifically, when $DBH$ was taken at the time of coring,  
At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used.
(see [issue #19 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/19))
If before, ... 
If after...
For all outside-in reconstructions, if a negative $DBH$ was predicted...

In either case we need bark thickness--ideally allometries describing the relationship between DBH and bark thickness. 
This is especially critical for thick-barked species
When bark thickness data were available, we generated allometries ... 
lognormal model with intercept forced to zero: lm(bark_depth.mm ~ -1 + log(dbh_no_bark.cm+1):bark_species, data = bark)
([issue #8 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/8))

## Appendix S2. Methods for comparing 
(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)

@@

\newpage 
![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png)
